Co-evolution of lexical meaning \& pragmatic use

### 2018-01-15 paper/cogsci-submission2.tex
first full draft for revision and revision letter

\begin{abstract}
 According to standard linguistic theory, the meaning of an utterance is the product of
  conventional semantic meaning and general pragmatic rules on language use. To investigate how
  cultural evolution of language plays out under this picture of the semantics-pragmatics
  interface, we present a game-theoretic model of the competition between types of language
  users, each endowed with a selection of lexical representations and a particular pragmatic
  disposition to act on them. Our model traces two evolutionary forces and their interaction:
  (i) pressure towards communicative efficiency and (ii) learning biases during the transfer of
  linguistic knowledge. We illustrate the model based on a case study on scalar
  implicatures. \rvv{The relationship between underspecified semantics and pragmatic inference that this study suggests is one of co-evolution. The maintenance of such semantics depends on pragmatic reasoning for their enrichment. In the opposite direction, complex semantics that can do away with mutual reasoning do not necessarily foster pragmatic reasoning and may even be encumbered by it.}
\end{abstract}

@@

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved; why it would be so pervasive across natural
languages; and why it is that some expressions systematically draw from it while others semantically conventionalize. Models of language evolution abound. There are simulation-based models studying
populations of communicating agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, many coming from game theory
\citep{Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,nowak:2006,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
of this work has focused on explaining basic properties such as compositionality and
combinatoriality
\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax},
but little attention has been paid to the interaction between conventional meaning and
pragmatic use. What is more, many mathematical models explain evolved meaning as a regularity
in the overt behavior of agents, abstracting from complex interactions between semantic
representations and pragmatic use. In contrast, we will here look at language users with a
richer linguistic make-up. More precisely, We spell out a model of the co-evolution of conventional meaning and pragmatic reasoning.
The objects of replication and selection are pairs consisting of a set of lexical
meanings and a manner of pragmatic behavior, which together define agents' linguistic choices. Pragmatic behaviors are represented using probabilistic models of pragmatic language use \citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. This
approach allows us to study the interaction between (i) pressure towards communicative
efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
such as inductive learning biases and sparse learning data \citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,KirbyGriffith2014:Iterated-Learni}. Considering the effects that the transmission of
linguistic knowledge through (iterated) learning have is important because neither semantic meanings nor pragmatic usage
patterns are directly observable. Instead, language learners have to infer these unobservables -- lexical meanings and manners of pragmatic behavior --
from the observable behavior in which they result -- overt linguistic choice in a context of interaction. We formalize this process as a form of
Bayesian inference. Our approach thereby contains a well-understood model of
iterated Bayesian learning \citep{griffiths+kalish:2005,griffiths+kalish:2007}, but combines it with functional
selection, here formalized as the most versatile dynamic from evolutionary game theory, the
replicator dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces
this model.

Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. We
discuss a setting in which the majority view of underspecified lexical meanings and systematic pragmatic
enrichments emerges if selection and transmission infidelity are combined. In particular, we
show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
may play a role in preventing the lexicalization of scalar inferences and thereby lead to the emergence of Gricean-like
pragmatic reasoning types. The results of this case study are critically assessed in the light of
the assumptions that feed our model in Section~\ref{sec:discussion}.

@@

\section{A model of co-evolving lexical representations and pragmatic behavior}
\label{sec:model}

@@

The idea that language is an adaptation to serve a communicative function is fundamental to
many synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of word
frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another, as natural languages need to be learnable to survive their
faithful transmission across generations. Furthermore, the effects of even small learning biases implicit in acquisition can build up and have quite striking impact on an evolving language in a process of iterated
learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

While natural languages are pressured for both communicative efficiency and learnability, these forces may
pull in opposite directions \citep[\S7]{christiansen+chater:2008}. Their opposition becomes particularly clear when considering the
extreme (\citealt{kemp+regier:2012,kirby+etal:2015}). A language consisting of a single
form-meaning association is easy to learn but lacking in expressivity. Conversely, a language
that lexicalizes a distinct form for a large number of different meanings is highly expressive
but challenging to acquire.

\subsection{The replicator mutator dynamic}
An elegant formal approach to capture the interaction between communicative efficiency and learnability is
the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. \rvv{The RMD is a population dynamic, which is to say that it describes change in the composition of a community made up of different types of agents. The population vector $\vec{x}$ codifies proportions or frequencies of types in a population, with $x_i$ being the frequency of type $i$. This population is infinite; its carrying capacity is held constant, $\sum\nolimits_i x_i = 1$. This allows us to track change that does not depend on varying population sizes nor their growth rates (but see, e.g., \citealt{nowak:2006} and \citealt[\S 5]{Skyrms2010:Signals} for alternatives that consider, among others, population size and its growth).} In its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$ in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
$i$ before the update step, (ii) the fitness $f_i$ of each type $i$ before the update, and
(iii) the probability $Q_{ji}$ that an agent who observes the behavior of 
an agent with type $j$ ends up acquiring type $i$:
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_h x_h f_h}\,.
\end{align}
The RMD consists of two components: fitness-based selection and transmission biases encoded in the so-called mutation matrix $Q$. This
becomes most transparent when we consider an equivalent formulation in terms of a step-wise
application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$
and its subsequent multiplication with a mutation matrix $Q$:
\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
\end{align}
where
\begin{align*}
      \left ( \text{RD}(\vec{x}) \right )_i 
         = \frac{x_i f_i}{\sum_k x_k f_k}
 \ \ \ \ \text{and} \ \ \ \ 
  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i\,.
\end{align*}
%

If the transmission matrix $Q$ is trivial in the sense that $Q_{ji}=1$ whenever $j=i$, the
dynamic reduces to the replicator dynamic. This situation is one where types are always replicated faithfully. The replicator dynamic is a model of fitness-based selection in which the relative frequency of type $i$ will increase with a gradient proportional to its average fitness in the population. In other words, selection comes into play whenever two or more types replicate at different rates due to differences in their fitness. This dynamic is popular and versatile because it can be derived from many abstract processes of biological and cultural
transmission and selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}, including conditional imitation
\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If fitness
$f_i$ is the same for all types $i$, the replicator step is the identity map
$ \left ( \text{RD}(\vec{x}) \right )_i = x_i$. In words, no difference in fitness translates into a lack of functional pressure: the population's type composition stays the same after the replicator step. In such cases the dynamic reduces to a process of iteration of the transmission bias encoded in $Q$: the rate by which one type changes into another after a generational turnover. As detailed below, in this way, the process in (\ref{eq:RMD_discrete}), and its equivalent in (\ref{eq:RMD_discrete_recast}), can contains a model of iterated learning \citep{griffiths+kalish:2005,griffiths+kalish:2007}.

@@

\begin{example}
  Consider a simple and abstract coordination game. Agents are of two types: positive or negative. If
  agents of different type interact with each other, they obtain a payoff of $0$. If negative meets
  negative, each receives a payoff of $1$. If positive meets positive, they get a payoff of $2$.  A population state is completely characterized by the proportion $x$ of negatives. The fitness
  of negatives in population state $x$ is $f_n(x) = x$, that of positives is $f_p(x) = 2-2x$.
  The average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation,
  the replicator dynamic will update $x$ to $\text{RD}(x) = \nicefrac{f(x)x}{\Phi(x)} = \nicefrac{x^2}{\Phi(x)}$. The update
  function $\text{RD}(x)$ of the replicator step is plotted in Fig.~\ref{fig:Updates_RMS} as the
  blue line. Rest points, for which $\text{RD}(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$.
  The former are attractors as nearby points converge to them. Points near $x=\nicefrac{2}{3}$
  also move towards 0 or 1. This is schematically pictured in the topmost phase portrait in
  Fig.~\ref{fig:Phase_RD}. Adding mutation changes the dynamic and its rest points. Let us assume that $Q_{ji} = .9$
  when $j=i$. This is the proportion of types that are replicated faithfully. Conversely, a proportion of $.1$ will change their type from positive to negative, and vice-versa. The update effect of mutation on its own are described by $\text{M}(x) = .9 x + .1 (1-x) = .8x + .1$, plotted as the linear green line in Fig.~\ref{fig:Updates_RMS}. As shown in Fig.~\ref{fig:Phase_RD}, in this example mutation alone has only one stable rest point. It is located at $x = .5$. If we first take the replicator step and
  then the mutation step in sequence, we obtain the replicator mutator dynamic
  $\text{RMD}(x) = \text{M}(\text{RD}(x)) = \nicefrac{.9x^2 - .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in
  Fig.~\ref{fig:Updates_RMS}. The rest points are at $x=.121$, $x=.903$ and $x=.609$. The
  former two are attractors (see Fig.~\ref{fig:Phase_RD}).
\end{example}

@@

Moving beyond an abstract example, our goal is to apply the RMD to investigate the co-evolution
of lexical representations and pragmatic behavior. To do so, we need to fix three things: (i) what the
relevant types are that populate $x$, (ii) how fitness derives from communicative success and (iii) how the
mutation matrix $Q$ is computed. These issues are addressed, one by one, in the following.

\subsubsection{Types: Lexica and pragmatic strategies}
\label{sec:languages+use}
Types are what evolution operates on. They define an agent's fitness, usually through a payoff
accrued in single interactions with other agents. Often types can be identified as the possible
acts in a game, e.g., either cooperating or defecting in a prisoner's dilemma. In other cases,
they may be thought of as general properties of an agent that influence her fitness, such as
being positive or negative in our previous example (whatever that means). For our present
purposes, types are identified by specific assumptions about their cognitive
make-up. More concretely, since we are interested in the evolutionary competition between different lexical representations and ways of using them in communication, a type is here defined as a pair consisting of a lexicon and a pragmatic strategy of language use.

Agents play signaling games, in which the speaker wants to communicate a world state $s$ with a
message $m$ to a hearer who receives $m$ but does not know $s$
\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
(possibly fuzzy) set of states. A pragmatic behavior specifies a probabilistic speaker
rule (a probabilistic choice of message for each state) and a probabilistic hearer rule (a
probabilistic choice of state for each message) given a lexicon. There are many ways of making these general notions more concrete. Broadly speaking, there are models of pragmatic language use coming from a game-theoretic tradition \citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and ones from a Bayesian tradition \citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}. We will not focus on the particular design choices that make up each approach \citep{qing+franke:2015}. Instead, we will stress how the approach, common to all of these models, of pragmatic language use as {\em rational language use} can be embedded within the replicator mutator dynamic to answer evolutionary questions at the semantics-pragmatics interface. Here is what we will assume in the remainder of this paper.\footnote{Readers more familiar with the Bayesian tradition should note that the definitions below follow the game-theoretic tradition in assuming rational sender choice to consist in sending messages that maximize expected utility, rather than message informativity (see \citealt{qing+franke:2015} for detailed discussion).}

Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}.  Lexica for this case would assign a
Boolean truth value, either $0$ for false or $1$ for true, to each state-message pair. The
following two lexica are minimal examples for the distinction between a lexicalized upper-bound
for \emph{some} in $\Lbound$ and the widely assumed logical semantics with only a lower-bound
in $\Llack$.

@@

Pragmatic strategies define dispositions to produce and interpret messages given a lexicon. We
distinguish between two kinds of pragmatic strategies. {\em Literal interlocutors} produce and
interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
rational language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
level $0$, corresponds to literal language use, as in Equations~(\ref{h:level0}) and
(\ref{s:level0}). Pragmatic language users of level $n + 1$ act (approximately) rational with
respect to level-$n$ behavior of their interlocutors, as in Equations~(\ref{h:leveln}) and
(\ref{s:leveln}).
\begin{alignat}{3}
&H_{0}(s \mid m;L) &&\propto pr(s) L_{[s,m]} \label{h:level0}\\
&S_{0}(m \mid s;L) &&\propto \exp(\lambda \; L_{[s,m]}) \label{s:level0}\\
&H_{n+1}(s|m;L) &&\propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) &&\propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
\end{alignat}

@@

In words, an agent's linguistic behavior -- what message to send when / which message to interpret how -- is defined by (i) her reasoning level and (ii) her lexicon. In what follows, this is what we identify an agent's type with in order to analyze under which conditions combinations of (i) and (ii) emerge through cultural evolution. Intuitively, literal behavior (level-$0$ reasoning) results from unreflected language use. Such agents produce and comprehend solely based on the truth-conditions of their lexicon. In a Gricean spirit, pragmatic behavior (level-$n+1$ reasoning) results from mutual reasoning about rational language use. Such agents produce and comprehend by reasoning about their interlocutor's likely behavior interpretation/production patterns. This reasoning is based on a conjecture resulting from their own behavior if they were in the role of their interlocutor. Higher order behavior is therefore, even if in tendency rational, quite simple. Crucially, we do not assume that agents know each other's types when interacting (e.g., what lexicon the interlocutor is using).

@@

In words, expected utility $\text{EU}(t_i,t_j)$ quantifies how successful communication between agents of types $i$ and $j$ is, with each type being a combination of a lexicon and a reasoning level (Section~\ref{sec:language+use}). Fitness $f_i$ indicates how well type $i$ fares in a population given what other types are found in it: it is type $i$'s weighted expected utility.

>>>>

\begin{abstract}
 According to standard linguistic theory, the meaning of an utterance is the product of
  conventional semantic meaning and general pragmatic rules on language use. To investigate how
  cultural evolution of language plays out under this picture of the semantics-pragmatics
  interface, we present a game-theoretic model of the competition between types of language
  users, each endowed with a selection of lexical representations and a particular pragmatic
  disposition to act on them. Our model traces two evolutionary forces and their interaction:
  (i) pressure towards communicative efficiency and (ii) transmission perturbations during the acquisition of
  linguistic knowledge. We illustrate the model based on a case study on scalar
  implicatures, which suggests the relationship between underspecified semantics and pragmatic inference to be one of co-evolution. The maintenance of underspecified semantics depends on pragmatic reasoning for their enrichment. In the opposite direction, complex semantics that can do away with mutual reasoning may not foster pragmatic reasoning, and may even be encumbered by it.
\end{abstract}

@@

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved; why it would be so pervasive across natural
languages; and why it is that some expressions systematically draw from it while others semantically conventionalize.

Models of language evolution abound. There are simulation-based models studying
populations of communicating agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, many coming from game theory
\citep{Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,nowak:2006,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
of this work has focused on explaining basic properties such as compositionality and
combinatoriality
\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax},
but little attention has been paid to the interaction between conventional meaning and
pragmatic use. What is more, many mathematical models explain evolved meaning as a regularity
in the overt behavior of agents, abstracting from complex interactions between semantic
representations and pragmatic use. In contrast, we will here look at language users with a
richer linguistic make-up. More precisely, we spell out a model of the evolution of conventional meaning and pragmatic reasoning.
The objects of replication and selection are pairs consisting of a set of lexical
meanings and a manner of pragmatic behavior, which together define agents' linguistic choices. Pragmatic behaviors are represented using probabilistic models of pragmatic language use \citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. This
approach allows us to study the interaction between (i) pressure towards communicative
efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
such as inductive learning biases and sparse learning data \citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,KirbyGriffith2014:Iterated-Learni}. Considering the effects that the transmission of
linguistic knowledge through (iterated) learning have is important because neither semantic meaning nor pragmatic usage
patterns are directly observable. Instead, language learners have to infer these unobservables from the observable behavior in which they result. We formalize this process as a form of
Bayesian inference. Our approach thereby contains a well-understood model of
iterated Bayesian learning \citep{griffiths+kalish:2005,griffiths+kalish:2007}, but combines it with functional
selection, here formalized as the most versatile dynamic from evolutionary game theory, the
replicator dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces
this model.

Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. We
discuss a setting in which the majority view of underspecified lexical meanings and systematic pragmatic
enrichments emerges if selection and transmission infidelity are combined. In particular, we
show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
may play a role in preventing the lexicalization of scalar inferences and thereby lead to the emergence of Gricean-like
pragmatic reasoning types. The results of this study are critically assessed in the light of
the assumptions that feed our model in Section~\ref{sec:discussion}.

@@

\section{A model of evolving lexical representations and pragmatic behavior}
\label{sec:model}

@@

The idea that language is an adaptation to serve a communicative function is fundamental to
many synchronic and diachronic analyses, at least since \citeposs{zipf:1949} explanation of word
frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another, as natural languages need to be learnable to survive their
faithful transmission across generations. Furthermore, the effects of even small learning biases implicit in acquisition can build up and have quite striking impact on an evolving language in a process of iterated
learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

While natural languages are pressured for both communicative efficiency and learnability, these forces may
pull in opposite directions \citep[\S7]{christiansen+chater:2008}. Their opposition becomes particularly clear when considering the
extreme (\citealt{kemp+regier:2012,kirby+etal:2015}). A language consisting of a single
form-meaning association is easy to learn but likely bad for communication with peers. Conversely, a language
that lexicalizes distinct forms for a large number of different meanings may be better for communication 
but challenging to acquire.

\subsection{The replicator mutator dynamic}
An elegant formal approach to capture the interaction between communicative efficiency and learnability is
the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. The RMD is a population dynamic, which is to say that it describes change in the composition of a community made up of different types of agents. The population vector $\vec{x}$ codifies proportions or frequencies of types in a population, with $x_i$ being the frequency of type $i$. This population is infinite; its carrying capacity is held constant, $\sum\nolimits_i x_i = 1$. This allows us to track change that does not depend on varying population sizes nor their growth rates (but see, e.g., \citealt{nowak:2006} and \citealt[\S 5]{Skyrms2010:Signals} for alternatives that consider, among others, population size and growth). In its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$ in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
$i$ before the update step, (ii) the fitness $f_i$ of each type $i$ before the update, and
(iii) the probability $Q_{ji}$ that an agent who observes the behavior of 
an agent with type $j$ ends up acquiring type $i$:
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_h x_h f_h}\,.
\end{align}
The RMD consists of two components: fitness-based selection and transmission biases encoded in the so-called mutation matrix $Q$. This
becomes most transparent when we consider an equivalent formulation in terms of a step-wise
application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$ and its subsequent multiplication with the mutation matrix $Q$:
\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
\end{align}
where
\begin{align*}
      \left ( \text{RD}(\vec{x}) \right )_i 
         = \frac{x_i f_i}{\sum_k x_k f_k}
 \ \ \ \ \text{and} \ \ \ \ 
  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i\,.
\end{align*}
%

If the transmission matrix $Q$ is trivial in the sense that $Q_{ji}=1$ whenever $j=i$, the
dynamic reduces to the replicator dynamic. This situation is one where types are always replicated faithfully. The replicator dynamic is a model of fitness-based selection in which the relative frequency of type $i$ will increase with a gradient proportional to its average fitness in the population. In other words, selection comes into play whenever two or more types replicate at different rates due to differences in their fitness. This dynamic is popular and versatile because it can be derived from many abstract processes of biological and cultural
transmission and selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}, including conditional imitation
\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If fitness
$f_i$ is the same for all types $i$, the replicator step is the identity map
$ \left ( \text{RD}(\vec{x}) \right )_i = x_i$. In words, no difference in fitness translates into a lack of functional pressure: the population's type composition stays the same after the replicator step. In such cases the dynamic reduces to a process of iteration of the transmission bias encoded in $Q$: the rate by which one type changes into another after a generational turnover. As detailed below, in this way, the process in (\ref{eq:RMD_discrete}), and its equivalent in (\ref{eq:RMD_discrete_recast}), can contain a model of iterated learning \citep{griffiths+kalish:2005,griffiths+kalish:2007}.

@@

\begin{example}
  Consider a simple and abstract coordination game. Agents are of two types: positive or negative. If
  agents of different type interact with each other, they obtain a payoff of $0$. If negative meets
  negative, each receives a payoff of $1$. If positive meets positive, they get a payoff of $2$.  A population state is completely characterized by the proportion $x$ of negatives. The fitness
  of negatives in population state $x$ is $f_n(x) = x$, that of positives is $f_p(x) = 2-2x$.
  The average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation,
  the replicator dynamic will update $x$ to $\text{RD}(x) = \nicefrac{f(x)x}{\Phi(x)} = \nicefrac{x^2}{\Phi(x)}$. The update
  function $\text{RD}(x)$ of the replicator step is plotted in Fig.~\ref{fig:Updates_RMS} as the
  blue line. Rest points, for which $\text{RD}(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$.
  The former are attractors; nearby points converge to them. Points near $x=\nicefrac{2}{3}$
  also move towards 0 or 1. This is schematically pictured in the topmost phase portrait in
  Fig.~\ref{fig:Phase_RD}. Adding mutation changes the dynamic and its rest points. Let us assume that $Q_{ji} = .9$
  when $j=i$. This is the proportion of types that are replicated faithfully. Conversely, a proportion of $.1$ will change their type from positive to negative, and vice-versa. The update effect of mutation on its own are described by $\text{M}(x) = .9 x + .1 (1-x) = .8x + .1$, plotted as the linear green line in Fig.~\ref{fig:Updates_RMS}. As shown in Fig.~\ref{fig:Phase_RD}, in this example mutation alone has only one stable rest point. It is located at $x = .5$. If we first take the replicator step and
  then the mutation step in sequence, we obtain the replicator mutator dynamic
  $\text{RMD}(x) = \text{M}(\text{RD}(x)) = \nicefrac{.9x^2 - .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in
  Fig.~\ref{fig:Updates_RMS}. The rest points are at $x=.121$, $x=.903$ and $x=.609$. The
  former two are attractors (see Fig.~\ref{fig:Phase_RD}).
\end{example}

@@

Moving beyond an abstract example, our goal is to apply the RMD to investigate the evolution
of lexical representations and pragmatic behavior. To do so, we need to fix three things: (i) what the
relevant types are that populate $x$, (ii) how fitness derives from communicative success and (iii) how the
mutation matrix $Q$ is computed. These issues are addressed, one by one, in the following.

\subsubsection{Types: Lexica and pragmatic strategies}
\label{sec:languages+use}
Types are what evolution operates on. They define an agent's fitness, usually through a payoff
accrued in single interactions with other agents. Often types can be identified as the possible
acts in a game, e.g., either cooperating or defecting in a prisoner's dilemma. In other cases,
they may be thought of as general properties of an agent that influence her fitness, such as
being positive or negative in our previous example (whatever that means). For our present
purposes, types are identified by specific assumptions about their cognitive
make-up. More concretely, since we are interested in the evolutionary competition between different lexical representations and ways of deploying them in communication, a type is here defined as a pair consisting of a lexicon and a linguistic strategy of language use.

Agents play signaling games, in which the speaker wants to communicate a world state $s$ with a
message $m$ to a hearer who receives $m$ but does not know $s$
\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
(possibly fuzzy) set of states. A linguistic behavior specifies a probabilistic speaker
rule (a probabilistic choice of message for each state) and a probabilistic hearer rule (a
probabilistic choice of state for each message) given a lexicon. There are many ways of making these general notions concrete. There are models of rational language use coming from a game-theoretic tradition \citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and ones from a Bayesian tradition \citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}. We will not focus on the particular design choices that make up each approach (on this see \citealt{qing+franke:2015}). Instead, we will stress how the approach, common to all of these models, of pragmatic language use as {\em rational language use} can be embedded within the replicator mutator dynamic to answer evolutionary questions at the semantics-pragmatics interface. Here is what we will assume in the remainder of this paper.\footnote{Readers more familiar with the Bayesian tradition than the game-theoretic one should note that the definitions below follow the latter in assuming rational sender choice to consist in sending messages that maximize expected utility, rather than message informativity (see \citealt{qing+franke:2015} for discussion).}

Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}.  Lexica for this case would assign a
Boolean truth value, either $0$ for false or $1$ for true, to each state-message pair. The
following two lexica are minimal examples for the distinction between a lexicalized upper-bound
for \emph{some}, in $\Lbound$, and the widely assumed logical semantics with only a lower-bound,
in $\Llack$.

@@

Linguistic strategies define dispositions to produce and interpret messages given a lexicon. We
distinguish between two kinds strategies. {\em Literal interlocutors} produce and
interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
rational language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
level $0$, corresponds to literal language use, as in Equations~(\ref{h:level0}) and
(\ref{s:level0}). Pragmatic language users of level $n + 1$ act (approximately) rational with
respect to level-$n$ behavior of their interlocutors, as in Equations~(\ref{h:leveln}) and
(\ref{s:leveln}).
\begin{alignat}{3}
&H_{0}(s \mid m;L) &&\propto pr(s) L_{[s,m]} \label{h:level0}\\
&S_{0}(m \mid s;L) &&\propto \exp(\lambda \; L_{[s,m]}) \label{s:level0}\\
&H_{n+1}(s|m;L) &&\propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) &&\propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
\end{alignat}

@@

In words, an agent's linguistic behavior -- what message to send when or which message to interpret how -- is defined by (i) her reasoning level and (ii) her lexicon. In what follows, this is what we identify an agent's type with, in order to analyze under which conditions combinations of (i) and (ii) emerge through cultural evolution. Intuitively, literal behavior (level-$0$ reasoning) results from unreflected language use. Such agents produce and comprehend solely based on the truth-conditions of their lexicon. In a Gricean spirit, pragmatic behavior (level-$n+1$ reasoning) results from mutual reasoning about rational language use. Such agents produce and comprehend by reasoning about their interlocutor's likely interpretation/production patterns. Higher order behavior is therefore, even if in tendency rational, quite simple. It does not assume agents to know each other's types when interacting (e.g., what lexicon the interlocutor is using); they simply act based on a conjecture of their own behavior if they were in the role of their interlocutor.

@@

In words, expected utility $\text{EU}(t_i,t_j)$ quantifies how successful communication between agents of types $i$ and $j$ is, with each type's behavior resulting from a combination of a lexicon and a reasoning level (Section~\ref{sec:languages+use}). Fitness $f_i$ indicates how well type $i$ fares in a population given what other types are found in it: it is type $i$'s weighted expected utility.
